A Philosophy of (Inductive) Testing

### 2024-04-10 mss.tex
Final version for proof.

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  20–23. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{A Philosophy of (Inductive) Testing}
\fancyfoot[LE,RO]{\thepage}

@@

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{1}

@@

Use what is useful and don't get hung-up on a particular framework or methodology. Record tests as you develop. You may think this slows down your development. That's not necessarily a bad thing. You will move forward with confidence that the last thing you worked on really works and you will find yourself thinking through your specification at a deeper level. This is important in a world of mostly incomplete and nebulous specs. What I am driving at is a cousin to test-driven development.\footnote{\citet{Beck2023} summarizes the principles of \textsc{tdd}.} Think of it as real-time test development.

@@

Descending from the hypothetical to the practical, every practical programming language comes with a unit testing environment. Hoon supplies \texttt{/lib/test} on the \texttt{\%base} desk, paired with the \texttt{/ted/test} thread; the \texttt{/lib} file should be copied to your development desk.\footnote{\citet{Davis2022} discusses this procedure in detail.} The rest of this article assumes a working knowledge of \texttt{/lib/test}.

@@

Under the hood, the Obelisk engine calls out to the urQL parser to create AST commands from the user-created scripts. The \texttt{AS OF} clause overrides the time passed in the bowl. Finally we weld together two results and check them against what we expect. The first check, expect-eq, is of the metadata returned by the last command. The second check is the final database state after the evolution.

Unit testing purism insists on specifically tailoring each test to one specific case of potential failure. This approach tends to be pedantic and frequently we pragmatically sneak in multiple independent tests in one test bundle, as in Listing~\ref{lst:bundle}.  \texttt{++test-alter-index-1} is not an atomic unit test but tests several qualities together. Values like whitespace and mixed-case labels are tedious to exhaustively test and well-suited to property based testing software like \texttt{\%quiz} (see below).

@@

We were intentionally dismissive of testing methodologies a\-bove because they are mere conventions. There is however some actual theory we can leverage in figuring out what and how to test.\footnote{\citet{WikipediaInduction} provides a modestly technical exposition of mathematical induction. You can skip it because we will explain it non-rigorously for lay programmers. We also quibble about the inclusion of zero as a natural number. While old school maths started the natural numbers with “1”, but computer science has since infected maths. Have you ever seen zero of anything? No—it is not natural at all. Zero is a very, very special number. It's an abstraction that does not map to anything in the physical world.}

@@

Input argument interactions are not the only source of combinatorial explosion in testing. Imagine that your function (gate) is a black box. You start submitting random input to figure out what the underlying algorithm is. However, it turns out that you discover that there is not one consistent algorithm for all inputs. Some values or ranges of values behave differently from others. (Unexpected UTF-8 whitespace characters are notorious for revealing head-scratching bugs—so imagine all possible inputs over complex XML.) You could model this behavior as multiple inductive types making for an even bigger cartesian explosion of inputs to test. It is no longer practical to construct all the tests required by our theory. The number is still finite, but impractically large. What is to be done?

@@

An automated testing solution called property-based testing may be applicable in these cases. The idea of property-based testing is to develop and instantiate invariant properties of the code and let software generate random inputs. The software runs the random inputs and tests the outputs against the invariant properties. An architecture for this approach was originally developed for Haskell and has since ported to many other languages, including Hoon.  \texttt{\%quiz} is a well-documented Hoon implementation \citetrepo{Hjort2023}.  Once again, this solution requires some deep thinking about the code and its specification.

@@

Lastly, whenever you fix a bona fide production bug (or one that a framework like \texttt{\%quiz} discovered), add a test case to address that circumstance. This not only provides the standard for when the bug has been fixed, but protects against regressing to the prior behavior (thus, a “regression test”). Congratulations, you have just refined your specification.

@@

In our experience failure modes are the most overlooked part of sofware development. It starts with passing insufficient, or no information from the programmed points of failure. So even before testing failure modes make sure you distinguish (i.e. make unique) each message from every point of failure and include any and all relevant information available for debugging.

\begin{lstlisting}[style=listingcode]
~|("cannot add duplicate key: {<row-key>}" !!)
\end{lstlisting}

The standard \texttt{/lib/test} library, as of this writing, can test for failure but not for an expected message.  A modified testing arm (Listing~\ref{lst:error}) can be included in a testing thread to facilitate this kind of testing.  Listing~\ref{lst:urql} shows a test for an expected error message in the Obelisk database engine which uses this functionality to verify that the correct error message is raised on crash.

>>>>

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  59–73. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{A Philosophy of (Inductive) Testing}
\fancyfoot[LE,RO]{\thepage}

@@

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{59}

@@

Use what is useful and don't get hung-up on a particular framework or methodology. Record tests as you develop. You may think this slows down your development. That's not necessarily a bad thing. You will move forward with confidence that the last thing you worked on really works and you will find yourself thinking through your specification at a deeper level. This is important in a world of mostly incomplete and nebulous specs. What I am driving at is a cousin to test-driven development.\footnote{\citet{Beck2023} summarizes the principles of {\sc tdd}.} Think of it as real-time test development.

@@

Descending from the hypothetical to the practical, every practical programming language comes with a unit testing environment. Hoon supplies \lstinline[style=inlinecode]{/lib/test} on the \lstinline[style=inlinecode]{%base} desk, paired with the \lstinline[style=inlinecode]{/ted/test} thread; the \texttt{/lib} file should be copied to your development desk.\footnote{\citet{Davis2022} discusses this procedure in detail.} The rest of this article assumes a working knowledge of \lstinline[style=inlinecode]{/lib/test}.

@@

Under the hood, the Obelisk engine calls out to the urQL parser to create {\sc ast} commands from the user-created scripts. The \texttt{AS OF} clause overrides the time passed in the bowl. Finally we weld together two results and check them against what we expect. The first check, \lstinline[style=inlinecode]{++expect-eq}, is of the metadata returned by the last command. The second check is the final database state after the evolution.

Unit testing purism insists on specifically tailoring each test to one specific case of potential failure. This approach tends to be pedantic and frequently we pragmatically sneak in multiple independent tests in one test bundle, as in Listing~\ref{lst:bundle}.  \lstinline[style=inlinecode]{++test-alter-index-1} is not an atomic unit test but tests several qualities together. Values like whitespace and mixed-case labels are tedious to exhaustively test and well-suited to property based testing software like \lstinline[style=inlinecode]{%quiz} (see below).

@@

We were intentionally dismissive of testing methodologies a\-bove because they are mere conventions. There is however some actual theory we can leverage in figuring out what and how to test.\footnote{\citeauthor{WikipediaInduction} (2024b) provides a modestly technical exposition of mathematical induction. You can skip it because we will explain it non-rigorously for lay programmers.  We also quibble about the inclusion of zero as a natural number.  While old school maths started the natural numbers with “1”, but computer science has since infected maths. (Have you ever seen zero of anything?   No—it is not natural at all.)  Zero is a very, very special number.  It's an abstraction that does not map to anything in the physical world.}

@@

Input argument interactions are not the only source of combinatorial explosion in testing. Imagine that your function (gate) is a black box. You start submitting random input to figure out what the underlying algorithm is. However, it turns out that you discover that there is not one consistent algorithm for all inputs. Some values or ranges of values behave differently from others. (Unexpected {\sc utf}-8 whitespace characters are notorious for revealing head-scratching bugs—so imagine all possible inputs over complex {\sc xml}.) You could model this behavior as multiple inductive types making for an even bigger cartesian explosion of inputs to test. It is no longer practical to construct all the tests required by our theory. The number is still finite, but impractically large. What is to be done?

@@

An automated testing solution called property-based testing may be applicable in these cases. The idea of property-based testing is to develop and instantiate invariant properties of the code and let software generate random inputs. The software runs the random inputs and tests the outputs against the invariant properties. An architecture for this approach was originally developed for Haskell and has since ported to many other languages, including Hoon.  \lstinline[style=inlinecode]{%quiz} is a well-documented Hoon implementation \citetrepo{Hjort2023}.  Once again, this solution requires some deep thinking about the code and its specification.

@@

Lastly, whenever you fix a bona fide production bug (or one that a framework like \lstinline[style=inlinecode]{%quiz} discovered), add a test case to address that circumstance. This not only provides the standard for when the bug has been fixed, but protects against regressing to the prior behavior (thus, a ``regression test''). Congratulations, you have just refined your specification.

@@

In our experience failure modes are the most overlooked part of sofware development. It starts with passing insufficient, or no information from the programmed points of failure. So even before testing failure modes make sure you distinguish (i.e. make unique) each message from every point of failure and include any and all relevant information available for debugging (cf. Listing~\ref{lst:dup}).

\begin{lstlisting}[
  float,
  style=listingcode,
  caption={Error message for duplicate key},
  label={lst:dup}
]
~|("cannot add duplicate key: {<row-key>}" !!)
\end{lstlisting}

The standard \lstinline[style=inlinecode]{/lib/test} library, as of this writing, can test for failure but not for an expected message.  A modified testing arm (Listing~\ref{lst:error}) can be included in a testing thread to facilitate this kind of testing.  Listing~\ref{lst:urql} shows a test for an expected error message in the Obelisk database engine which uses this functionality to verify that the correct error message is raised on crash.
